---
title: 'Data Analysis: Pokemon, DND, and NYC Gifted Dataset'
author: "W. Durosier"
date: "2025-03-09"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This project is to provide a comprehensive analysis involving data cleaning, transformation, and visualization for three different datasets: Pokémon competitive analysis, Dungeons & Dragons (DND) characters, and NYC Gifted and Talented Grades

# Load packages
```{r}
library(readr)
library(tidyr)
library(dplyr)
library(janitor)
library(tidyverse)
library(stringr)
library(readxl)
library(ggplot2)
```

## Creating path for the Datasets


```{r data} 

white.csv <-("C:\\Users\\wduro\\OneDrive - City University of New York\\DATA607\\pokemon_competitive_analysis.csv")

write_tsv <- ("C:\\Users\\wduro\\OneDrive - City University of New York\\DATA607\\dnd_chars_all.tsv.txt") 

white.csv <- ("C:\\Users\\wduro\\Downloads\\NYC Gifted and Talented Grades 2018-19 - Sheet5.csv")

```

# Readig the data 

The three datasets were loaded into R using read_csv and read_tsv functions.
Each dataset was read in with its specific file path.

```{r}
pokemon_data <- read_csv("C:\\Users\\wduro\\OneDrive - City University of New York\\DATA607\\pokemon_competitive_analysis.csv", show_col_types = FALSE)

dnd_data <- read_tsv ("C:\\Users\\wduro\\OneDrive - City University of New York\\DATA607\\dnd_chars_all.tsv.txt", show_col_types = FALSE)

 nyc_gifted_data <- read_csv("C:\\Users\\wduro\\Downloads\\NYC Gifted and Talented Grades 2018-19 - Sheet5.csv", show_col_types = FALSE)
```

```{r name-duplicate}

colnames(nyc_gifted_data)

colnames(nyc_gifted_data) <- make.names(colnames(nyc_gifted_data), unique = TRUE)

head(nyc_gifted_data)

nyc_gifted_data <- nyc_gifted_data[!duplicated(colnames(nyc_gifted_data))]
```

```{r datasets}

head(pokemon_data)

head(dnd_data)

head(nyc_gifted_data)

```
## Inspecting and Cleaning the Data:


Pokemon Dataset:
Missing values were handled using drop_na().
Created a new variable total_stats as a sum of multiple existing statistics.

# Check column names and missing values for pokemon_data

```{r}

colnames(pokemon_data)
missing_pokemon <- colSums(is.na(pokemon_data))
missing_pokemon

```
## Handdling missing Values

#Dropping rows with missing values to disply simplysity,then checking the data cleanliness.
```{r data-pokemon}

pokemon_data_cleaned <- pokemon_data %>%
  drop_na()

head(pokemon_data_cleaned)

```
# cleaning the datasets

```{r clean-pokemon}

pokemon_data_cleaned <- pokemon_data %>%
  drop_na()


head(pokemon_data_cleaned)

```

#DND Dataset:

Checked for missing values and removed rows with NAs.
```{r}

colnames(dnd_data)
missing_dnd <- colSums(is.na(dnd_data))
missing_dnd

```
```{r}

dnd_data_cleaned <- dnd_data %>%
  drop_na()


head(dnd_data_cleaned)

```

#NYC Gifted Dataset:

Converted the Entering Grade Level column into a factor variable and dropped any rows with missing values.
Used pivot_longer to reshape the dataset for better analysis, especially for the school type-related columns.

```{r}

colnames(nyc_gifted_data)
missing_nyc <- colSums(is.na(nyc_gifted_data))
missing_nyc

```


```{r}

nyc_gifted_data_cleaned <- nyc_gifted_data %>%
  drop_na() %>%
  mutate(grade = as.factor(`Entering.Grade.Level`))  


head(nyc_gifted_data_cleaned)

```


# Pivoting the data to Check the transformed long format


```{r}

nyc_long_data <- nyc_gifted_data_cleaned %>%
  pivot_longer(cols = starts_with("School"), 
               names_to = "school_type", 
               values_to = "school_value")


head(nyc_long_data)

```


# New column

Check the new data set for new column created


```{r}

pokemon_data_cleaned <- pokemon_data_cleaned %>%
  mutate(total_stats = attack + defense + speed + sp_atk + sp_def)


head(pokemon_data_cleaned)

```

# Summary statistics

 for NYC Gifted data, Pokemon data,  DND data, basic summary statistics were generated using the summary() function.

```{r}

summary(nyc_gifted_data_cleaned)


summary(pokemon_data_cleaned)


summary(dnd_data_cleaned)

```

## Visualizations:

#Pokemon Dataset:

A histogram was plotted to show the distribution of the total_stats variable.

```{r}
ggplot(pokemon_data_cleaned, aes(x = total_stats)) +
  geom_histogram(binwidth = 10, fill = "orange", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Total Stats for Pokémon", 
       x = "Total Stats", 
       y = "Frequency") +
  theme_minimal()

```



# Saving the cleaned data

The cleaned datasets were saved into new CSV files for further use or sharing

```{r}

write_csv(pokemon_data_cleaned, "C:\\Users\\wduro\\OneDrive - City University of New York\\DATA607\\pokemon_data_cleaned.csv")

write_csv(nyc_gifted_data_cleaned, "C:\\Users\\wduro\\OneDrive - City University of New York\\DATA607\\nyc_gifted_data_cleaned.csv")

write_csv(dnd_data_cleaned, "C:\\Users\\wduro\\OneDrive - City University of New York\\DATA607\\dnd_data_cleaned.csv")

```

#Accesing cleaned data to analyze

```{r}
pokemon_data <- read_csv ("C:\\Users\\wduro\\OneDrive - City University of New York\\DATA607\\pokemon_data_cleaned.csv")

head(pokemon_data)
```

```{r}
nyc_gifted_data <- read_csv ("C:\\Users\\wduro\\OneDrive - City University of New York\\DATA607\\nyc_gifted_data_cleaned.csv")

head(nyc_gifted_data)
```


```{r}
dnd_data <- read_csv("C:\\Users\\wduro\\OneDrive - City University of New York\\DATA607\\dnd_data_cleaned.csv")

head(dnd_data)
```
# Data Cleaning and Transformation:
 Ensuring data are tidy and transformed.
 
 
# Tidy the Pokemon dataset
```{r}

pokemon_data_cleaned <- pokemon_data %>%
  drop_na() %>% 
  mutate(total_stats = attack + defense + speed + sp_atk + sp_def)


head(pokemon_data_cleaned)

```
# DND Dataset
We'll check for missing values and remove rows with NA. Additionally, we can transform the dataset (e.g., for specific character attributes).
```{r}
dnd_data_cleaned <- dnd_data %>%
  drop_na()

head(dnd_data_cleaned)
```

# NYC Gifted Dataset
We'll ensure that the Entering Grade Level column is properly converted to a factor and handle any other transformations needed.

```{r}
nyc_gifted_data_cleaned <- nyc_gifted_data %>%
  drop_na() %>%
  mutate(grade = as.factor(`Entering.Grade.Level`))

head(nyc_gifted_data_cleaned)
```



## Conclusion, 
In this following project the goal was to transform and cleaned three data sets.
The project involved data cleaning, handling missing values, transforming variables, and performing some basic analysis and visualizations. The datasets were saved for future reference and additional analysis.

#Key Learnings:

Data Cleaning: You handled missing data, standardized column names, and reshaped data for easier analysis.

Transformation: You created new variables (e.g., total_stats) to enhance the analysis.

Visualizations: You generated clear and effective visualizations, particularly for the Pokémon dataset, to understand the distribution of key variables.

Tidy Data Principles: You applied key tidying techniques to ensure your datasets were in a "long" format when needed.

